Combined annotation exploration for SCPCP000006

### 2025-02-27 analyses/cell-type-wilms-tumor-06/07_combined_annotation_across_samples_exploration.Rmd
notebook after review

_Where `cnv.thr` and `pred.thr` need to be discussed_

| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`  | `VWF`| no CNV |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`  | `PTPRC`, `CD163`, `CD68`| no CNV |
| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells  | -  | -|

@@

@@

- the predicted compartment in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`

@@

```{r input}
# sample metadata
sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)

# the cytoband file that will be used to extract the size of each chromosome arm for plotting only
arm_order_file <- file.path(module_base, "results", "references", "hg38_cytoBand.txt")

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        # label transfer from the fetal kidney reference
        cell_type = srat$fetal_kidney_predicted.cell_type,
        compartment = srat$fetal_kidney_predicted.compartment,

        # predicted.scores from the label transfer from the fetal kidney reference
        cell_type.score = srat$fetal_kidney_predicted.cell_type.score,
        compartment.score = srat$fetal_kidney_predicted.compartment.score,

@@

        # marker genes
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts"),
        VIM = FetchData(object = srat, vars = "ENSG00000026025", layer = "counts"),
        CITED1 = FetchData(object = srat, vars = "ENSG00000125931", layer = "counts"),
        CDH1 = FetchData(object = srat, vars = "ENSG00000039068", layer = "counts"),
        PODXL = FetchData(object = srat, vars = "ENSG00000128567", layer = "counts"),
        COL6A3 = FetchData(object = srat, vars = "ENSG00000163359", layer = "counts"),
        SIX2 = FetchData(object = srat, vars = "ENSG00000170577", layer = "counts"),
        NCAM1 = FetchData(object = srat, vars = "ENSG00000149294", layer = "counts"),
        THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts"),

        # loss global estimation per chromosome
        has_loss_chr1p = srat$has_loss_chr1p,
        has_loss_chr2p = srat$has_loss_chr2p,
        has_loss_chr3p = srat$has_loss_chr3p,
        has_loss_chr4p = srat$has_loss_chr4p,
        has_loss_chr5p = srat$has_loss_chr5p,
        has_loss_chr6p = srat$has_loss_chr6p,
        has_loss_chr7p = srat$has_loss_chr7p,
        has_loss_chr8p = srat$has_loss_chr8p,
        has_loss_chr9p = srat$has_loss_chr9p,
        has_loss_chr10p = srat$has_loss_chr10p,
        has_loss_chr11p = srat$has_loss_chr11p,
        has_loss_chr12p = srat$has_loss_chr12p,

        has_loss_chr16p = srat$has_loss_chr16p,
        has_loss_chr17p = srat$has_loss_chr17p,
        has_loss_chr18p = srat$has_loss_chr18p,
        has_loss_chr19p = srat$has_loss_chr19p,
        has_loss_chr20p = srat$has_loss_chr20p,

        has_loss_chr1q = srat$has_loss_chr1q,
        has_loss_chr2q = srat$has_loss_chr2q,
        has_loss_chr3q = srat$has_loss_chr3q,
        has_loss_chr4q = srat$has_loss_chr4q,
        has_loss_chr5q = srat$has_loss_chr5q,
        has_loss_chr6q = srat$has_loss_chr6q,
        has_loss_chr7q = srat$has_loss_chr7q,
        has_loss_chr8q = srat$has_loss_chr8q,
        has_loss_chr9q = srat$has_loss_chr9q,
        has_loss_chr10q = srat$has_loss_chr10q,
        has_loss_chr11q = srat$has_loss_chr11q,
        has_loss_chr12q = srat$has_loss_chr12q,
        has_loss_chr13q = srat$has_loss_chr13q,
        has_loss_chr14q = srat$has_loss_chr14q,
        has_loss_chr15q = srat$has_loss_chr15q,
        has_loss_chr16q = srat$has_loss_chr16q,
        has_loss_chr17q = srat$has_loss_chr17q,
        has_loss_chr18q = srat$has_loss_chr18q,
        has_loss_chr19q = srat$has_loss_chr19q,
        has_loss_chr20q = srat$has_loss_chr20q,
        has_loss_chr21q = srat$has_loss_chr21q,
        has_loss_chr22q = srat$has_loss_chr22q,

        # dupli global estimation per chromosome
        has_dupli_chr1p = srat$has_dupli_chr1p,
        has_dupli_chr2p = srat$has_dupli_chr2p,
        has_dupli_chr3p = srat$has_dupli_chr3p,
        has_dupli_chr4p = srat$has_dupli_chr4p,
        has_dupli_chr5p = srat$has_dupli_chr5p,
        has_dupli_chr6p = srat$has_dupli_chr6p,
        has_dupli_chr7p = srat$has_dupli_chr7p,
        has_dupli_chr8p = srat$has_dupli_chr8p,
        has_dupli_chr9p = srat$has_dupli_chr9p,
        has_dupli_chr10p = srat$has_dupli_chr10p,
        has_dupli_chr11p = srat$has_dupli_chr11p,
        has_dupli_chr12p = srat$has_dupli_chr12p,

        has_dupli_chr16p = srat$has_dupli_chr16p,
        has_dupli_chr17p = srat$has_dupli_chr17p,
        has_dupli_chr18p = srat$has_dupli_chr18p,
        has_dupli_chr19p = srat$has_dupli_chr19p,
        has_dupli_chr20p = srat$has_dupli_chr20p,

        has_dupli_chr1q = srat$has_dupli_chr1q,
        has_dupli_chr2q = srat$has_dupli_chr2q,
        has_dupli_chr3q = srat$has_dupli_chr3q,
        has_dupli_chr4q = srat$has_dupli_chr4q,
        has_dupli_chr5q = srat$has_dupli_chr5q,
        has_dupli_chr6q = srat$has_dupli_chr6q,
        has_dupli_chr7q = srat$has_dupli_chr7q,
        has_dupli_chr8q = srat$has_dupli_chr8q,
        has_dupli_chr9q = srat$has_dupli_chr9q,
        has_dupli_chr10q = srat$has_dupli_chr10q,
        has_dupli_chr11q = srat$has_dupli_chr11q,
        has_dupli_chr12q = srat$has_dupli_chr12q,
        has_dupli_chr13q = srat$has_dupli_chr13q,
        has_dupli_chr14q = srat$has_dupli_chr14q,
        has_dupli_chr15q = srat$has_dupli_chr15q,
        has_dupli_chr16q = srat$has_dupli_chr16q,
        has_dupli_chr17q = srat$has_dupli_chr17q,
        has_dupli_chr18q = srat$has_dupli_chr18q,
        has_dupli_chr19q = srat$has_dupli_chr19q,
        has_dupli_chr20q = srat$has_dupli_chr20q,
        has_dupli_chr21q = srat$has_dupli_chr21q,
        has_dupli_chr22q = srat$has_dupli_chr22q
      ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()
```

@@

@@

```{r fig.width=20, fig.height=15, out.width='100%', results='asis'}

do_BoxPlot_mean <- function(df, group.by, comparison){
  tmp <- df |>
    select(!!sym(group.by), sample_id, contains("has_loss_chr"), contains("has_dupli_chr")) |>
    group_by(!!sym(group.by), sample_id) |>
    summarise_all(.funs = mean) |>
    pivot_longer(cols = -c(sample_id,!!sym(group.by)) ) |>
    rename("chromosome_arm" = "name", "percentage" = "value") |>
    mutate(chromosome_arm = factor(chromosome_arm, levels = c(has_dupli_chr, has_loss_chr)))

@@

We simply checked for each chromosome if the cell `has_cnv_chr`.
Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `has_cnv_score` will be FALSE.
Would the cell have more than `r params$cnv_threshold_high`  chromosome with CNV, the global `has_cnv_score` will be TRUE.
The default is set to `unknown`.

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold_high ~ "CNV",
    has_cnv_score <= params$cnv_threshold_low ~ "no CNV",
    .default = "unknown"
  ))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & 
      has_cnv_score == "no CNV" ~ "normal",

    compartment %in% c("normal", "immune", "endothelium") &
      cell_type.score > params$predicted.celltype.threshold ~ "normal",

    compartment %in% c("fetal_nephron", "stroma") & 
      has_cnv_score == "CNV" ~ "cancer",

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
`r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))

DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))

@@

cell_type_df$first.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",

    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",

    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("B cell",
                                    "CD4-positive, alpha-beta T cell",
                                    "conventional dendritic cell",
                                    "lymphocyte",
                                    "macrophage",
                                    "mast cell",
                                    "monocyte",
                                    "natural killer cell",
                                    "neutrophil",
                                    "plasmacytoid dendritic cell") ~ "immune",

    # assign cancer cells based on condition
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
    .default = "unknown"
  ))

cell_type_df$second.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 1, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
cell_type_df_sub <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
ggplot(cell_type_df_sub, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 1, shape = 19) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells;
and `subclonal` (transparent) when detected in 20 to 70% of the cells.

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

# Here we define the vector of ordered chromosomes arms that will be used for plotting

has_dupli_chr <- c("has_dupli_chr1p",                                      
                   "has_dupli_chr1q",

                   "has_dupli_chr2p",
                   "has_dupli_chr2q",

                   "has_dupli_chr3p",
                   "has_dupli_chr3q",

                   "has_dupli_chr4p",
                   "has_dupli_chr4q",

                   "has_dupli_chr5p",
                   "has_dupli_chr5q",

                   "has_dupli_chr6p",
                   "has_dupli_chr6q",

                   "has_dupli_chr7p",
                   "has_dupli_chr7q",

                   "has_dupli_chr8p",
                   "has_dupli_chr8q",

                   "has_dupli_chr9p",
                   "has_dupli_chr9q",

                   "has_dupli_chr10p",
                   "has_dupli_chr10q",

                   "has_dupli_chr11p",
                   "has_dupli_chr11q",

                   "has_dupli_chr12p",
                   "has_dupli_chr12q",

                   "has_dupli_chr13p",
                   "has_dupli_chr13q",

                   "has_dupli_chr14p",
                   "has_dupli_chr14q",

                   "has_dupli_chr15p",
                   "has_dupli_chr15q",

                   "has_dupli_chr16p",
                   "has_dupli_chr16q",

                   "has_dupli_chr17p",
                   "has_dupli_chr17q",

                   "has_dupli_chr18p",
                   "has_dupli_chr18q",

                   "has_dupli_chr19p",
                   "has_dupli_chr19q",

                   "has_dupli_chr20p",
                   "has_dupli_chr20q",

                   "has_dupli_chr21p",
                   "has_dupli_chr21q",

                   "has_dupli_chr22p",
                   "has_dupli_chr22q"
                   )

#################################################################################

# define the vector of ordered chromosomes arms##################################
has_loss_chr <-  c("has_loss_chr1p",                                        
                   "has_loss_chr1q",

                   "has_loss_chr2p",
                   "has_loss_chr2q",

                   "has_loss_chr3p",
                   "has_loss_chr3q",

                   "has_loss_chr4p",
                   "has_loss_chr4q",

                   "has_loss_chr5p",
                   "has_loss_chr5q",

                   "has_loss_chr6p",
                   "has_loss_chr6q",

                   "has_loss_chr7p",
                   "has_loss_chr7q",

                   "has_loss_chr8p",
                   "has_loss_chr8q",

                   "has_loss_chr9p",
                   "has_loss_chr9q",

                   "has_loss_chr10p",
                   "has_loss_chr10q",

                   "has_loss_chr11p",
                   "has_loss_chr11q",

                   "has_loss_chr12p",
                   "has_loss_chr12q",

                   "has_loss_chr13p",
                   "has_loss_chr13q",

                   "has_loss_chr14p",
                   "has_loss_chr14q",

                   "has_loss_chr15p",
                   "has_loss_chr15q",

                   "has_loss_chr16p",
                   "has_loss_chr16q",

                   "has_loss_chr17p",
                   "has_loss_chr17q",

                   "has_loss_chr18p",
                   "has_loss_chr18q",

                   "has_loss_chr19p",
                   "has_loss_chr19q",

                   "has_loss_chr20p",
                   "has_loss_chr20q",

                   "has_loss_chr21p",
                   "has_loss_chr21q",

                   "has_loss_chr22p",
                   "has_loss_chr22q"
                   )

#################################################################################
# Here we define the size of each chromosome arm that will be used for plotting of the mean CNV profile

  # Load cytoBand file
  cytoBand <- read.table(gzfile(arm_order_file), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
  colnames(cytoBand) <- c("chrom", "start", "end", "band", "stain")

  # Extract chromosome arm information
  cytoBand <- cytoBand |>
    mutate(arm = substr(band, 1, 1))  # Extract 'p' or 'q' from the band column

  # Define arm boundaries
  chromosome_arms <- cytoBand |>
    group_by(chrom, arm) |>
    summarize(
      Start = min(start),
      End = max(end),
      .groups = "drop"
    ) |>
    mutate(Size = End - Start,
           dupli = paste0("has_dupli_", chrom, arm),
           loss = paste0("has_loss_", chrom, arm)) |> 
  filter(arm != "")

  chromosome_arms <- chromosome_arms |>
    mutate(Size = Size / sum(Size))

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

# prepare the table for the number of sample with a (sub)clonal loss

CNV_profile_loss <- cell_type_df_sub |>
  select(sample_id, contains("has_loss_chr")) |>
  group_by(sample_id) |>
  summarise_all(.funs = mean) |>
  pivot_longer(cols = -sample_id) |>
  rename("chromosome_arm" = "name", "percentage" = "value") |>
  left_join( chromosome_arms, by = c("chromosome_arm"="loss")) |>
  mutate(chromosome_arm = factor(chromosome_arm, levels = has_loss_chr)) |>
  mutate(absolute_70 = case_when((percentage > 0.7) ~ -1,
         .default = 0),
         absolute_10 = case_when((percentage > 0.1) ~ -1,
         .default = 0))

@@

CNV_profile_gain <- cell_type_df_sub |>
  select(sample_id, contains("has_dupli_chr")) |>
  group_by(sample_id) |>
  summarise_all(.funs = mean) |>
  pivot_longer(cols = -sample_id) |>
  rename("chromosome_arm" = "name", "percentage" = "value") |>
  left_join( chromosome_arms, by = c("chromosome_arm"="dupli")) |>
  mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
  mutate(absolute_70 = case_when((percentage > 0.7) ~ 1,
         .default = 0),
         absolute_10 = case_when((percentage > 0.1) ~ 1,
         .default = 0))

# plot

loss_profile <- ggplot(CNV_profile_loss, aes(x = chromosome_arm, y = absolute_10, width = 20 * Size)) +
  geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#41b6c4") +
  scale_x_discrete(labels = gsub("has_loss_", "", levels(droplevels(CNV_profile_loss$chromosome_arm))))  +
  theme_classic() +
  ylab("Number of occurrence") +
  xlab("chromosom arms") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))

gain_profile <- ggplot(CNV_profile_gain, aes(chromosome_arm, y = absolute_10, width = 20* Size)) +
  geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#ce1256") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        text = element_text(size = 15)) +
  ylab(" ")

>>>>

| first level annotation | second level annotation | selection of the cells                                                                        | marker genes for validation | CNV validation                             |
| ---------------------- | ----------------------- | --------------------------------------------------------------------------------------------- | --------------------------- | ------------------------------------------ |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr`                                   | `VWF`                       | no CNV                                     |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr`                                        | `PTPRC`, `CD163`, `CD68`    | no CNV                                     |
| normal                 | kidney                  | `compartment == "fetal_nephron" & predicted.score > pred.thr & cnv_score < cnv.thr.lower`     | `CDH1`, `PODXL`, `LTL`      | no CNV                                     |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr.upper`            | `VIM`                       | no CNV                                     |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr.upper`                                         | `VIM`                       | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CITED1`                    | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr.upper`| `CDH1`                      | `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells                                                                         | -                           | -                                          |

with the defined parameters:

- `pred.thr` = `r params$predicted.celltype.threshold`

- `cnv.thr.lower` = `r params$cnv_threshold_low`

- `cnv.thr.upper` = `r params$cnv_threshold_high`

@@

```{r }
# sample metadata
sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")
metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)
```

@@

- the predicted cell type in `fetal_kidney_predicted.cell_type` and `fetal_kidney_predicted.cell_type.score`

@@

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        srat@meta.data,

@@

        # marker genes
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts"),
        VIM = FetchData(object = srat, vars = "ENSG00000026025", layer = "counts"),
        CITED1 = FetchData(object = srat, vars = "ENSG00000125931", layer = "counts"),
        CDH1 = FetchData(object = srat, vars = "ENSG00000039068", layer = "counts"),
        PODXL = FetchData(object = srat, vars = "ENSG00000128567", layer = "counts"),
        COL6A3 = FetchData(object = srat, vars = "ENSG00000163359", layer = "counts"),
        SIX2 = FetchData(object = srat, vars = "ENSG00000170577", layer = "counts"),
        NCAM1 = FetchData(object = srat, vars = "ENSG00000149294", layer = "counts"),
        THY1 = FetchData(object = srat, vars = "ENSG00000154096", layer = "counts")

      ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }

  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()
```

@@

```{r input}
# the cytoband file that will be used to extract the size of each chromosome arm for plotting only
arm_order_file <- file.path(module_base, "results", "references", "hg38_cytoBand.txt")
# Load cytoBand file
cytoBand <- read.table(gzfile(arm_order_file), header = FALSE, sep = "\t", stringsAsFactors = FALSE)
colnames(cytoBand) <- c("chrom", "start", "end", "band", "stain")
# Here we define the size of each chromosome arm that will be used for plotting of the mean CNV profile
  # Extract chromosome arm information
  cytoBand <- cytoBand |>
    mutate(arm = substr(band, 1, 1))  # Extract 'p' or 'q' from the band column

  # Define arm boundaries
  chromosome_arms <- cytoBand |>
    group_by(chrom, arm) |>
    summarize(
      Start = min(start),
      End = max(end),
      .groups = "drop"
    ) |>
    mutate(Size = End - Start,
           dupli = paste0("has_dupli_", chrom, arm),
           loss = paste0("has_loss_", chrom, arm),
           chrom_arm = glue::glue("{chrom}{arm}")) |> 
    mutate(Size = Size / sum(Size)) |>
    # Define chromosome arm order
    mutate(chrom_arm = factor(chrom_arm, levels = c(paste0("chr", rep(1:22, each = 2), c("p", "q")),
                                                  "chrXp", "chrXq", "chrYp", "chrYq"))) |>
    # Sort genes by Chromosome arm and Start position
    arrange(chrom_arm, Start)  |> 
    filter(arm != "")

@@

```{r fig.width=20, fig.height=15, out.width='100%', results='asis'}
do_BoxPlot_mean <- function(df, group.by, comparison){
  tmp <- df |>
    select(!!sym(group.by), sample_id, contains("has_loss_chr"), contains("has_dupli_chr")) |>
    group_by(!!sym(group.by), sample_id) |>
    summarise_all(.funs = mean) |>
    pivot_longer(cols = -c(sample_id,!!sym(group.by)) ) |>
    rename("chromosome_arm" = "name", "percentage" = "value") |>
    mutate(chromosome_arm = factor(chromosome_arm, levels = c(has_dupli_chr, has_loss_chr)))

@@

We simply checked for each chromosome the cell `cnv_chr`.
Would the cell have less than `r params$cnv_threshold_low` chromosome with CNV, the global `cnv_score` will be FALSE.
Would the cell have more than `r params$cnv_threshold_high`  chromosome with CNV, the global `cnv_score` will be TRUE.
The default is set to `unknown`.

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(cnv_score = rowSums(cell_type_df[, grepl("has_loss_chr|has_dupli_chr", colnames(cell_type_df))], na.rm = TRUE)) |>
  mutate(cnv_score = case_when(
    cnv_score > params$cnv_threshold_high ~ "CNV",
    cnv_score <= params$cnv_threshold_low ~ "no CNV",
    .default = "unknown"
  ))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    fetal_kidney_predicted.compartment %in% c("fetal_nephron", "stroma") & 
    cnv_score == "no CNV" ~ "normal",

    fetal_kidney_predicted.compartment %in% c("normal", "immune", "endothelium") &
    fetal_kidney_predicted.compartment.score > params$predicted.celltype.threshold ~ "normal",

    fetal_kidney_predicted.compartment %in% c("fetal_nephron", "stroma") & 
    cnv_score == "CNV" ~ "cancer",

@@

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain _unknown_

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation)) +
  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))

cell_type_df |>
  dplyr::count(first.level_annotation, sample_id) |>
  DT::datatable()

@@

cell_type_df$first.level_annotation[cell_type_df$sample_id %in% none_reference_samples] <- "unknown"

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
      cell_type_df$fetal_kidney_predicted.compartment   == c("fetal_nephron") &
      cell_type_df$cnv_score                            == "no CNV" ~ "kidney",

      cell_type_df$fetal_kidney_predicted.compartment   == c("stroma") &
      cell_type_df$cnv_score                            == "no CNV" ~ "normal stroma",

      cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
      cell_type_df$fetal_kidney_predicted.cell_type     == c("endothelial cell") ~ "endothelium",

      cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
      cell_type_df$fetal_kidney_predicted.cell_type    %in% c("B cell",
                                    "CD4-positive, alpha-beta T cell",
                                    "conventional dendritic cell",
                                    "lymphocyte",
                                    "macrophage",
                                    "mast cell",
                                    "monocyte",
                                    "natural killer cell",
                                    "neutrophil",
                                    "plasmacytoid dendritic cell") ~ "immune",

    # assign cancer cells based on condition
      cell_type_df$fetal_kidney_predicted.compartment    == c("stroma") &
      cell_type_df$cnv_score                             == "CNV" ~ "cancer stroma",

      cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
      cell_type_df$cnv_score                             == "CNV" &
      cell_type_df$fetal_kidney_predicted.cell_type      == "mesenchymal cell" ~ "blastema",

      cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
      cell_type_df$cnv_score                             == "CNV" &
      cell_type_df$fetal_kidney_predicted.cell_type      != "mesenchymal cell" ~ "cancer epithelium",

      sample_id                                         %in% none_reference_samples ~ "unknown",

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
cell_type_df_sub <- cell_type_df[cell_type_df$second.level_annotation != "unknown",]
ggplot(cell_type_df_sub, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation)) +
  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

For a sample, a CNV is defined as `clonal` (opaque) if detected in more than 70% of the cells;
and `subclonal` (transparent) when detected in at least 10% of the cells.

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
# Here we define the vector of ordered chromosomes arms that will be used for plotting
has_dupli_chr <- names(cell_type_df)[stringr::str_starts(names(cell_type_df), "has_dupli_chr")]

# define the vector of ordered chromosomes arms
has_loss_chr <- names(cell_type_df)[stringr::str_starts(names(cell_type_df), "has_loss_chr")]

@@

```{r fig.height=10, fig.width=20, message=FALSE, warning=FALSE, out.width='100%', results='asis'}
# prepare the table for the number of sample with a (sub)clonal loss

CNV_profile_loss <- cell_type_df_sub |>
                    select(sample_id, contains("has_loss_chr")) |>
                    group_by(sample_id) |>
                    summarise_all(.funs = mean) |>
                    pivot_longer(
                      cols = -sample_id, 
                      names_to = "chromosome_arm", 
                      values_to = "percentage") |>
                    left_join(chromosome_arms, by = c("chromosome_arm"="loss")) |>
                    mutate(chromosome_arm = factor(chromosome_arm, levels = has_loss_chr)) |>
                    mutate(
                      absolute_70 = ifelse(percentage > 0.7, -1, 0), 
                      absolute_10 = ifelse(percentage > 0.1, -1, 0))

@@

CNV_profile_gain <- cell_type_df_sub |>
                    select(sample_id, contains("has_dupli_chr")) |>
                    group_by(sample_id) |>
                    summarise_all(.funs = mean) |>
                    pivot_longer(
                      cols = -sample_id, 
                      names_to = "chromosome_arm", 
                      values_to = "percentage") |>
                    left_join(chromosome_arms, by = c("chromosome_arm"="dupli")) |>
                    mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
                    mutate(
                      absolute_70 = ifelse(percentage > 0.7, 1, 0), 
                      absolute_10 = ifelse(percentage > 0.1, 1, 0))

# plot
# Here we chose to loop on chromosome arms to then be able to scale each bar to the length of the corresponding genomic region using wrap_plot. 
# Of note, the resulting plot is way better than the one obtained with facet_wrap or facet_grid so far.

loss_profile <- list()
for(chr in unique(CNV_profile_loss$chromosome_arm)){
  loss_profile[[chr]] <- ggplot(CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr,], 
                                aes(x = chromosome_arm, y = absolute_10)) +
                        geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
                        geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#41b6c4") +
                        scale_x_discrete(labels = gsub("has_loss_", "", chr))  +
                        theme_void() +
                        ylim(c(-40,0)) +
                        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
}
loss_profile <- wrap_plots(loss_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$loss %in% unique(CNV_profile_loss$chromosome_arm)], guides = "collect")
